Sperm should evolve to make female meiosis fair.

### 2014-07-14 writeup_and_notes/male_dependent_drive.tex
first go at revisions by YB next do some math and sims for female vs male suppression of female drive

%Why do animals allow male control of female meiosis?
%Why does female meiosis allow the opportunity for exploitation by selfish sperm?
%sperm and egg genomes both favour the peaceful resolution of female meiotic drive
%strange bed fellows

%\title{Why does female meiosis allow the opportunity for exploitation by selfish sperm? BETTER TITLE?}
\title{
%Co-operation between the sexes trumps selfish control of female  meiotic drive. 
%Or 
Sperm should evolve to make female meiosis fair.
%Sperm do not evolve to collaborate in female meiotic drive. 
%Or Why does female meiosis allow the opportunity for exploitation by selfish sperm? 
}

\author{Yaniv Brandvain$^1$ \\ email: ybrandvain@gmail.com  \and Graham Coop$^2$ \\ email: gmcoop@ucdavis.edu }

@@

%FIgures 
{\bf Abstract:}
Genomic conflicts arise when an allele gains an evolutionary advantage at a cost to organismal fitness. 
%Because only one of four products of female meiosis is transmitted to offspring, 
O\"{o}genesis is inherently susceptible to such conflicts
because alleles compete to be the  product of female meiosis transmitted to the egg. 
%(meiotic drive).  
Alleles that distort meiosis in their favor (i.e. meiotic drivers) often decrease organismal fitness,
	and therefore indirectly favor the evolution of mechanisms to suppress meiotic drive. % drive the rapid evolution of suppression mechanisms, and
In this light, many facets of  o\"{o}genesis and gametogenesis have been 
	interpreted as mechanisms of protection against genomic outlaws. 
Why then is female meiosis often left uncompleted until after
fertilization in many animals -- 
	potentially providing an opportunity for sperm alleles to
        meddle with its outcome and help like-alleles drive in heterozygous females? %and facilitate meiotic drive? 
The population genetic theory presented herein suggests that
	sperm nearly always evolve to increase the fairness of female
        meiosis in the face of genomic conflicts. %%felt this
                                %%might be an important addition as
                                %%they will spread if no cost to organism
These results are consistent with current knowledge of sperm-dependent meiotic drivers 
(loci whose distortion of female meiosis depends on sperm genotype),
	and suggest that the requirement of fertilization for the
        completion of female meiosis
        %, and variation in stage of meiosis that
        %fertilization occurs at 
        potentially represents a mechanism employed by females 
	 to ensure a fair meiosis. %prevent meiotic drive.
\newpage

@@

It is also not the case that sperm is mechanistically incapable of influencing female meiosis.
In \emph{C. elegans}, premature deployment of the aster -- a vital component of mitotic
machinery provided by the sperm, disrupts MII meiotic segregation
in the egg, leading to a triploid zygote \citep{McNally2012}. 
Thus, there is direct mechanistic evidence that sperm can influence female meiosis.
 Additionally, genetic evidence from the \emph{In} and \emph{Om}  drive systems in mice suggests that the 
 transmission of alternate alleles in heterozygous females can depend on sperm haplotype.  
 While ruling out the alternative hypothesis of early selection on zygotes in these cases is challenging  \citep[see pages
52-54 in][]{Burt2006}, it appears that the extent to which \emph{In} and \emph{Om} distort the second female meiotic division partially depends on the genotype of the fertilizing sperm \citep{Agulnik1993,Wu2005}.  
% Both of these female drive systems operate by distorting the second meiotic division, 
% and in both systems the outcome of female meiosis depends the genotype of the fertilizing sperm \citep{Agulnik1993,Wu2005}.  
%While such genetic evidence is suggestive, we note the difficulty in
%ruling out the alternative hypothesis of early selection on zygotes \citep[pages
%52-54][]{Burt2006}.

We develop population genetic models exploring the evolution of sperm influence on female meiotic drive. %, focussing on models in which `self-promoting' alleles in sperm facilitate drive of like genotypes during gametogenesis in heterozygous females. 
We first focus on models in which `self-promoting' alleles in
  sperm facilitate drive of like alleles during gametogenesis in heterozygous females. 
%In this article we explore through simple population genetic models the consequences of alleles that influence the outcome of female meiosis.
These models show that such sperm-acting alleles 
	have more difficulty invading a population than do traditional meiotic drivers, 
	 and under most circumstances, cannot be maintained as a balanced polymorphism.
Because self-promoting drivers are unlikely to create a sustained genomic conflict, 
	female meiosis will have little incentive to evolve resistance to them.
We then examine models in which a novel sperm-acting allele modifies the efficacy of a polymorphic meiotic driver. 
Such models universally reveal that sperm-acting drive modifiers are favored only if they suppress drive. 
These results highlight the fact that the interests of sperm and maternal  genomes' are often aligned, as both are invested in the fate of the resultant zygote (as was speculated for the \emph{In} locus \citep{Pomiankowski1993}).
Therefore, there is little selective benefit to females in preventing sperm to influence female meioses.
Rather, features of o\"{o}genesis  may evolve to facilitate an interaction with sperm %with female meiosis
	as yet another mechanism to enforce an equitable transfer of genetic material.

@@

In the standard single-locus, biallelic model of female meiotic drive,
the driving allele is transmitted to the egg in heterozygotes with
probability  $d > 1/2$, regardless of sperm genotype \cite[e.g. ][ and see Model 1 in the Appendix for more details]{Ubeda2004}. %The timing of fertilization relative to female meiosis places another constraint on $d$, for example, if fertilization (and therefore, sperm dependent drive) takes place at MII (as in mammals),
%	female drive requires an uneven number of crossovers between the centromere and the drive locus, 
%	so $d$ is bounded to be $<0.83$ \citep[see ][ for discussion]{Buckler1999}.
	%YB - MOVED THIS TO THE APPENDIX
          To depict a case
of a self-promoting meiotic driver,  we modify this standard model such 
	that the driver is only effective when fertilized by a sperm
        carrying that allele (see Figure \ref{Eggsperm}A and Model 2
        in the Appendix and Supplementary Worksheet). 
We then identify the conditions allowing for the spread of this self-promoting driver, 
	and evaluate whether a driver of this form could generate a sustained conflict favoring the evolution of suppressors. 
We conclude our single locus results with an analysis of a related model (Model 3) - in which
drivers influence their transmission in females via paternal genotype,
rather than sperm haplotype.

For comparative purposes, we first briefly present the standard drive model %as Model 1 in the Appendix, 
%equations \ref{eqn.A1}-\ref{eqn.A3}, 
	%and refer readers to the previous literature on this model 
	\citep[see e.g. ][ for additional results]{Prout1973,Ubeda2004}. 
Assuming that the driving allele is deleterious in both sexes, but fully recessive 
	(i.e. the fitness of drive homozygotes equals $w_{BB}=1-s$ and other genotypic fitnesses equal $w_{AA}=w_{AB}=1$), 
	it always invades because,  when rare it occurs predominantly in heterozygotes and therefore drives without a fitness cost. 
However, when $s$ is large ($s>(2d-1)/(2)$, solid black line in Figure \ref{Eggsperm}B) a driver cannot fix and
will be maintained as a protected polymorphism \citep{Prout1973}. 
The parameter space where the allele can invade but not fix is shown in white
        in Figure \ref{Eggsperm}B. %, and it can fix in the colored regions. 
When the allele is maintained as a polymorphism, it provides an opportunity for the evolution of
	drive suppressors, corresponding well to empirical examples of
        female meiotic drive \citep[reviewed in ][]{Burt2006}.

In contrast to a traditional driver, which drives but pays effectively
	no fitness cost when rare, 
a self-promoting driver specifically creates low fitness drive homozygotes 
	by uniting driving female gametes with sperm enabling that drive.
        % zygotes when it drives. in heterozygotes. 
%, and therefore there is a drive-associated fitness cost even when rare. 
It must therefore overcome a drive-associated homozygous fitness cost simply to spread when rare. 
The conditions allowing the invasion of a self-promoting driver
 	are consequently far more restrictive than those for a standard meiotic driver.
When rare, a fully recessive, self-promoting driver can only invade when $s$ 
	%, the selection against the drive homozygote, 
	is less than approximately $(2 d - 1)/(4 d)$ -- see dashed black line in Figure \ref{Eggsperm}B. 
This analytical approximation, derived from Equation \eqref{deltadriver} assuming Hardy-Weinberg, 
	closely matches results obtained by exact numerical iteration (Figure \ref{Eggsperm}B).

@@

%%%sperm-acting facilitator
When the modifier of drive arises on the drive background (i.e. in coupling phase),
	is tightly linked to the driver, and enhances drive 
	we label  this non-recombining drive/modifier haplotype as the $B^+$ allele.  
The $B^+$ allele acts in sperm to increase the effectiveness of drive for both
	the $B$ and  $B^{+}$ alleles in $AB$ and $AB^{+}$
        heterozygotes (see Figure \ref{Eggsperm_3_allele_cartoon}A2, and
        Model 4 in the Appendix and Supplementary Worksheet). 
Naively, $B^{+}$ may spread by capitalizing on the additional drive it causes; however,  
	this is not the case for a few simple reasons. 
First, 
	the novel $B^{+}$ haplotype
	arises when the ancestral driver 
	is at drive-selection balance, 
	and therefore immediately suffers a homozygous fitness cost.  
Worse yet, a novel $B^{+}$ haplotype most often helps 
	the $B$  allele drive ($B^+$ sperm meeting $AB$ eggs), because $B$ is initially more common than $B^{+}$. 
Therefore, sperm-acting drive facilitator alleles experience a profound disadvantage 
	in this scenario, even more so than under the previous two allele model. 
We have found no parameter range of this
	three allele system that allows the sperm-acting drive facilitator $B^{+}$ to
	invade the population (Appendix Model 4, eqn. \eqref{coupling}, and Mathematical Supplement).

\begin{figure}
\begin{center}
	\includegraphics[width = 0.7
          \textwidth]{Figures/twolocus.eps} \end{center}
\caption{Models of a sperm-acting drive modifier tightly linked to a meiotic driver. 
{\bf (A) } 
Sperm carrying the derived allele at the modifier locus (filled squares) alters transmission  
	at the driving allele (filled circles) during female meiosis.
        Alleles at these two tightly linked loci form three haplotypes (top of A). 	
	{\bf{A1)}} In the standard model of drive there is no
        variation at the modifier, 
	and the driver is transmitted to the egg with probability $d_0$. 
	{\bf{A2)}} The modifier allele increases the transmission of the drive allele ($d_1 > d_0$), and due to their shared genetic background, also increases its drive. 
	{\bf{A3 \& A4)}} The sperm-acting modifier acts to decrease
        drive  ($d_1=1/2$ in A3 \& A4, or more generally, $d_1 < d_0$) and arises on the same or opposite
        background from the driver (A3 \& A4 respectively). 
%		In {\bf{A3}} the drive suppressor exists on the drive background, forming the $B^-$ haplotype. 
%		In {\bf{A4}} the drive suppressor exists on the non-drive background, forming the $A^-$ haplotype. 		
{\bf{(B)}} Invasion of a sperm-acting drive suppressor linked to a driver. 
	After the driver ($B$ haplotype) reaches drive selection equilibrium, we introduce a sperm acting drive modifier. 
	We assume full drive ($D_0 =1$), a recessive lethal fitness cost to drive ($h_s=0$, $s=1$) and that the sperm-acting modifier results in a fair meiosis. 
{\bf {B1)}} The $B^-$ allele replaces the ancestral drive haplotype, but segregates at a lower equilibrium frequency.
{\bf {B2)}} The $A^-$ allele replaces the ancestral non-driving haplotype, and in this case, removes the driver from the population. }
 \label{Eggsperm_3_allele_cartoon}
\end{figure}

%%%sperm-acting drive suppressors
While  sperm enhancement of a female drive cannot displace a polymorphic female driver, sperm based drive suppressors can. 
%	that is sperm-acting alleles that act to restore  fairness of female meiosis, can. %If these sperm-acting drive suppressors arise on
Imagine a sperm-acting allele that restores fairness to female meiosis arises on
	the drive background, creating a
        third allele $B^{-}$ (Figure
          \ref{Eggsperm_3_allele_cartoon}A3, Model 4 in Appendix). This new allele still
        experiences female drive when fertilized by A or B sperm, but 
       it does not drive when fertilized by another $B^{-}$ so it
       avoid the excess formation of low fitness homozygotes. This
     allows the $B^{-}$ to displace the ancestral driver (Figure
     \ref{Eggsperm_3_allele_cartoon}B1, Equation \ref{coupling}), 
	and often returns to a lower equilibrium frequency than the the $B$ allele, further decreasing the extent of drive in the population.  
If this sperm-acting drive suppressor arises on 
	the non-driving $A$ background ( i.e. in repulsion phase, 
	creating a third allele $A^{-}$, Figure \ref{Eggsperm_3_allele_cartoon}A4, Model 5), 
	or is unlinked to the drive locus (Model 6), it readily invades a population segregating
	for the drive system (Equations \ref{A-} and
        \ref{unlinked}).  %%DELETED FIG REF HERE.
This allele lowers the frequency of the original driver (perhaps to zero),
	and spreads to fixation if it does not carry strong fitness costs
	(Figure \ref{Eggsperm_3_allele_cartoon}B2). 
This result is consistent with previous work showing that drive suppressors unlinked to, or in repulsion phase with drivers usually invade polymorphic drive systems \citep[e.g. ][]{Brandvain2012}.  
Therefore, all two-locus models of sperm influence on female drive suggest that
	sperm will evolve to oppose female meiotic drive.

%More surprisingly, sperm-acting suppressors of female drive can spread
%	even when tightly linked to the original drive system (Figures \ref{Eggsperm_3_allele_cartoon} and \ref{Trajectories_of_supressors}B). 
%The $B^{-}$ haplotype (representing the tightly linked drive/drive suppressor complex) displaces the original driver 
%	by driving when not costly, but avoiding low fitness homozygotes by disrupting drive when transmitted to sperm. %  ($BB^-$ and $B^-B^-$). 
%Despite displacing the ancestral driver (the $B$ haplotype), the equilibrium frequency of the $B^-$ haplotype is often lower than that of $B$ \yb{(point to supp?)}.

%\gc{Do we mention the In and Om locus here or in the discussion as a
%	 potential e.g. Do we mention the fact that the frequency of these
%	 newly invade sperm dodging alleles can be lower than the original system?}\\

@@

It seems that allowing sperm to influence the outcome of female meiosis would generate a confluence of these potential conflicts -- 
	sperm could actually assist an allele that distorts female meiosis.
However, this is not the case.
We find that an allele which acts through sperm to distort female meiosis in its favor %YB [too awkward / unclear?] 
	can rarely spread through a population if it bears any cost. 
Additionally, when this self-promoting driver can spread, it can only rarely 
	be maintained as a protected polymorphism, and due to its positive frequency dependence,  
	it spends very little time at intermediate frequency.
As such, this type of exploitation cannot generate a sustained genetic conflict.
It is therefore unlikely
	that female o\"{o}genesis and meiosis will evolve to prevent their effect.  
This likely explains why females can delay the completion of meiosis until after fertilization 
	without risking exploitation by collaborations between female
        drivers and sperm alleles.
%Male modification of recombination rates. \yb{trying to think through this \dots{}}

Why is it that an allele that biases female meiosis in its favor can generate a genetic conflict, but an allele in sperm that assists this female driver cannot? 
So long as the transmission advantage of female meiotic drive outweigh the organismal fitness cost to heterozygotes, the female driver can spread when rare, and it increases in 		
	frequency until the fitness cost to homozygotes balances the transmission advantage.
By contrast, a sperm promoter of female drive is only effective when matched with a heterozygote female -- meaning that, when rare, this allele rarely enhances female drive. 
Even worse, when it does so it will preferentially find itself in a low fitness drive homozygote. 
Not only are drive-promoting sperm alleles unable to create a sustained genetic conflict, 
	but alleles in sperm with the opposite effect - that is those that prevent their own drive through female meiosis do maintain a polymorphism and 
	provide evolution  with time and opportunity to further minimize drive.
This is because such drive suppressing alleles reduced their chances
of forming low fitness homozygotes. 
%In fact, an allele in sperm which causes the alternate allele to drive can be favored, a counterintuitive result 
%	consistent with the tenuous evidence that the non-driving
%        $ln^-$ allele is over transmitted through female meiosis in $ln^-/ln^+$ heterozygote when fertilized by $ln^+$
%        sperm \cite{agulnick, Pomiankowski1993}.   
% YB - WE DID NOT TALK ABOUT THIS IN THE RESULTS [ALTHOUGH IT'S TRUE... PERHAPS TOO DISTRACTING]
More generally, natural selection favors alleles that act through sperm to
 reduce the opportunity of female meiotic drive regardless of linkage or phase.

@@

\section*{Appendix}
We highlight simple analytical results from: 
	the standard model of female meiotic drive (Model 1); a sperm-dependent 
	model of female meiotic drive (Model 2); a male-genotype-dependent model of
	female meiotic drive (Model 3); and more complex two-locus sperm-dependent drive models (Models 4-6). 
All models include a biallelic locus ($A/B$) with non-driving and driving alleles in frequencies $f_A$ and $f_B$, 
	respectively, while more complex models include more loci and/or alleles.  
Transmission rules describing the outcomes of all matings in each
	model are presented in a Supplementary Worksheet. 
The fitness of genotype, $g$ is sex-independent and equals $1$, $1-h_s$, and $1-s$ for genotypes $AA$, $AB$, and $BB$, respectively. %  and  $w_{AA} = 1$, $w_{AB} = 1-s_h$, and $w_{BB}=1-s$. 
Genotypic frequencies  equal  $f_g$ for adults in the current generation, 
	$f_g'$ in the next generation of zygotes (i.e. after recombination, random mating, drive, and syngamy, but before selection)
	and $f_g''$ in adults after selection. 
After a complete generation $f_g'' = f_g' w_g/\bar{w}$, where $\bar{w}$ is the population mean fitness and equals $\Sigma w_g f_g'$. 
Because many of our analyses assume genotype frequencies expected under Hardy Weinberg Equilibrium (HWE), we note which analytical results are approximations. 
We verify these approximations with exact numerical iterations in Supplementary Figures and Supplementary Scripts, and detail the derivation of these approximations in our Mathematical Supplement.

>>>>

\title{  Sperm should evolve to make female meiosis fair.  }

\author{Yaniv Brandvain$^1$ \and Graham Coop$^2$ }

@@

\newpage 
\flushleft{{\bf{Running title:}} Sperm prevent female meiotic drive} \\
{\bf{Keywords:}} Meiosis; Meiotic drive; Fertilization; Genomic conflict \\
{\bf{Corresponding author:}} Yaniv Brandvain. \emph{Address:} Department of Plant Biology, University of Minnesota --€“ Twin
Cities. St. Paul MN, 55108. \emph{Phone:} 612-624-4375 \emph{Email:} ybrandvain@gmail.com

{\bf Abstract:}
Genomic conflicts arise when an allele gains an evolutionary advantage at a cost to organismal fitness. 
O\"{o}genesis is inherently susceptible to such conflicts
because alleles compete to be the  product of female meiosis transmitted to the egg. 
Alleles that distort meiosis in their favor (i.e. meiotic drivers) often decrease organismal fitness,
	and therefore indirectly favor the evolution of mechanisms to suppress meiotic drive. 
In this light, many facets of  o\"{o}genesis and gametogenesis have been 
	interpreted as mechanisms of protection against genomic outlaws. 
Why then is female meiosis often left uncompleted until after
fertilization in many animals -- 
	potentially providing an opportunity for sperm alleles to
        meddle with its outcome and help like-alleles drive in heterozygous females? 
The population genetic theory presented herein suggests that
	sperm nearly always evolve to increase the fairness of female
        meiosis in the face of genomic conflicts. 
These results are consistent with current knowledge of sperm-dependent meiotic drivers 
(loci whose distortion of female meiosis depends on sperm genotype),
	and suggest that the requirement of fertilization for the
        completion of female meiosis
        potentially represents a mechanism employed by females 
	 to ensure a fair meiosis.
\newpage

@@

It is also not the case that sperm is mechanistically incapable of influencing female meiosis.
In \emph{C. elegans}, premature deployment of the aster -- a vital component of mitotic
machinery provided by the sperm, disrupts MII meiotic segregation
in the egg, leading to a triploid zygote \citep{McNally2012}. 
Thus, there is direct mechanistic evidence that sperm can influence female meiosis.
 Additionally, genetic evidence from the \emph{In} and \emph{Om}  drive systems in mice suggests that the 
 transmission of alternate alleles in heterozygous females can depend on sperm haplotype.  
 While ruling out the alternative hypothesis of early selection on zygotes in these cases is challenging  \citep[see pages
52-54 in][]{Burt2006}, it appears that the extent to which \emph{In} and \emph{Om} distort the second female meiotic division partially depends on the genotype of the fertilizing sperm \citep{Agulnik1993,Wu2005}.

We develop population genetic models exploring the evolution of sperm influence on female meiotic drive. 
We first focus on models in which `self-promoting' alleles in
  sperm facilitate drive of like alleles during gametogenesis in heterozygous females. 
These models show that such sperm-acting alleles 
	have more difficulty invading a population than do traditional meiotic drivers, 
	 and under most circumstances, cannot be maintained as a balanced polymorphism.
Because self-promoting drivers are unlikely to create a sustained genomic conflict, 
	female meiosis will have little incentive to evolve resistance to them.
We then examine models in which a novel sperm-acting allele modifies the efficacy of a polymorphic meiotic driver. 
Such models universally reveal that sperm-acting drive modifiers are favored only if they suppress drive. 
These results highlight the fact that the interests of sperm and maternal  genomes' are often aligned, as both are invested in the fate of the resultant zygote (as was speculated for the \emph{In} locus \citep{Pomiankowski1993}).
Therefore, there is little selective benefit to females in preventing sperm to influence female meioses.
Rather, features of o\"{o}genesis  may evolve to facilitate an interaction with sperm %with female meiosis
	as yet another mechanism to enforce an equitable transfer of genetic material.

@@

In the standard single-locus, biallelic model of female meiotic drive,
the driving allele is transmitted to the egg in heterozygotes with
probability  $d > 1/2$, regardless of sperm genotype \cite[e.g. ][ and see Model 1 in the Appendix for more details]{Ubeda2004}. 
 To depict a case of a self-promoting meiotic driver,  we modify this standard model such 
	that the driver is only effective when fertilized by a sperm
        carrying that allele (see Figure \ref{Eggsperm}A and Model 2
        in the Appendix and File S1 ). 
We then identify the conditions allowing for the spread of this self-promoting driver, 
	and evaluate whether a driver of this form could generate a sustained conflict favoring the evolution of suppressors. 
We conclude our single locus results with an analysis of a related model (Model 3) - in which
drivers influence their transmission in females via paternal genotype,
rather than sperm haplotype.

For comparative purposes, we first briefly present the standard drive model 
	\citep[see e.g. ][ for additional results]{Prout1973,Ubeda2004}. 
Assuming that the driving allele is deleterious in both sexes, but fully recessive 
	(i.e. the fitness of drive homozygotes equals $w_{BB}=1-s$ and other genotypic fitnesses equal $w_{AA}=w_{AB}=1$), 
	it always invades because,  when rare it occurs predominantly in heterozygotes and therefore drives without a fitness cost. 
However, when $s$ is large ($s>(2d-1)/(2)$, solid black line in Figure \ref{Eggsperm}B) a driver cannot fix and
will be maintained as a protected polymorphism \citep{Prout1973}. 
The parameter space where the allele can invade but not fix is shown in white
        in Figure \ref{Eggsperm}B. %, and it can fix in the colored regions. 
When the allele is maintained as a polymorphism, it provides an opportunity for the evolution of
	drive suppressors, corresponding well to empirical examples of
        female meiotic drive \citep[reviewed in ][]{Burt2006}.

In contrast to a traditional driver, which drives but pays effectively
	no fitness cost when rare, 
a self-promoting driver specifically creates low fitness drive homozygotes 
	by uniting driving female gametes with sperm enabling that drive.
It must therefore overcome a drive-associated homozygous fitness cost simply to spread when rare. 
The conditions allowing the invasion of a self-promoting driver
 	are consequently far more restrictive than those for a standard meiotic driver.
When rare, a fully recessive, self-promoting driver can only invade when $s$ 
	is less than approximately $(2 d - 1)/(4 d)$ -- see dashed black line in Figure \ref{Eggsperm}B. 
This analytical approximation, derived from Equation \eqref{deltadriver} assuming Hardy-Weinberg, 
	closely matches results obtained by exact numerical iteration (Figure \ref{Eggsperm}B).

@@

%%%sperm-acting facilitator
When the modifier of drive arises on the drive background (i.e. in coupling phase),
	is tightly linked to the driver, and enhances drive 
	we label  this non-recombining drive/modifier haplotype as the $B^+$ allele.  
The $B^+$ allele acts in sperm to increase the effectiveness of drive for both
	the $B$ and  $B^{+}$ alleles in $AB$ and $AB^{+}$
        heterozygotes (see Figure \ref{Eggsperm_3_allele_cartoon}A2, and
        Model 4 in the Appendix and File S1 ). 
Naively, $B^{+}$ may spread by capitalizing on the additional drive it causes; however,  
	this is not the case for a few simple reasons. 
First, 
	the novel $B^{+}$ haplotype
	arises when the ancestral driver 
	is at drive-selection balance, 
	and therefore immediately suffers a homozygous fitness cost.  
Worse yet, a novel $B^{+}$ haplotype most often helps 
	the $B$  allele drive ($B^+$ sperm meeting $AB$ eggs), because $B$ is initially more common than $B^{+}$. 
Therefore, sperm-acting drive facilitator alleles experience a profound disadvantage 
	in this scenario, even more so than under the previous two allele model. 
We have found no parameter range of this
	three allele system that allows the sperm-acting drive facilitator $B^{+}$ to
	invade the population (Appendix Model 4, eqn. \eqref{coupling}, and File S2  ).

\begin{figure}
\begin{center}
	\includegraphics[width = 0.7
          \textwidth]{Figures/twolocus.eps} \end{center}
\caption{Models of a sperm-acting drive modifier tightly linked to a meiotic driver. 
{\bf (A) } 
Sperm carrying the derived allele at the modifier locus (filled squares) alters transmission  
	at the driving allele (filled circles) during female meiosis.
        Alleles at these two tightly linked loci form three haplotypes (top of A). 	
	{\bf{A1)}} In the standard model of drive there is no
        variation at the modifier, 
	and the driver is transmitted to the egg with probability $d_0$. 
	{\bf{A2)}} The modifier allele increases the transmission of the drive allele ($d_1 > d_0$), and due to their shared genetic background, also increases its drive. 
	{\bf{A3 \& A4)}} The sperm-acting modifier acts to decrease
        drive  ($d_1=1/2$ in A3 \& A4, or more generally, $d_1 < d_0$) and arises on the same or opposite
        background from the driver (A3 \& A4 respectively). 
{\bf{(B)}} Invasion of a sperm-acting drive suppressor linked to a driver. 
	After the driver ($B$ haplotype) reaches drive selection equilibrium, we introduce a sperm acting drive modifier. 
	We assume full drive ($D_0 =1$), a recessive lethal fitness cost to drive ($h_s=0$, $s=1$) and that the sperm-acting modifier results in a fair meiosis. 
{\bf {B1)}} The $B^-$ allele replaces the ancestral drive haplotype, but segregates at a lower equilibrium frequency.
{\bf {B2)}} The $A^-$ allele replaces the ancestral non-driving haplotype, and in this case, removes the driver from the population. }
 \label{Eggsperm_3_allele_cartoon}
\end{figure}

%%%sperm-acting drive suppressors
While  sperm enhancement of a female drive cannot displace a polymorphic female driver, sperm based drive suppressors can. 
Imagine a sperm-acting allele that restores fairness to female meiosis arises on
	the drive background, creating a
        third allele $B^{-}$ (Figure
          \ref{Eggsperm_3_allele_cartoon}A3, Model 4 in Appendix). This new allele still
        experiences female drive when fertilized by A or B sperm, but 
       it does not drive when fertilized by another $B^{-}$ so it
       avoid the excess formation of low fitness homozygotes. This
     allows the $B^{-}$ to displace the ancestral driver (Figure
     \ref{Eggsperm_3_allele_cartoon}B1, Equation \ref{coupling}), 
	and often returns to a lower equilibrium frequency than the the $B$ allele, further decreasing the extent of drive in the population.  
If this sperm-acting drive suppressor arises on 
	the non-driving $A$ background ( i.e. in repulsion phase, 
	creating a third allele $A^{-}$, Figure \ref{Eggsperm_3_allele_cartoon}A4, Model 5), 
	or is unlinked to the drive locus (Model 6), it readily invades a population segregating
	for the drive system (Equations \ref{A-} and
        \ref{unlinked}).  %%DELETED FIG REF HERE.
This allele lowers the frequency of the original driver (perhaps to zero),
	and spreads to fixation if it does not carry strong fitness costs
	(Figure \ref{Eggsperm_3_allele_cartoon}B2). 
This result is consistent with previous work showing that drive suppressors unlinked to, or in repulsion phase with drivers usually invade polymorphic drive systems \citep[e.g. ][]{Brandvain2012}.  
Therefore, all two-locus models of sperm influence on female drive suggest that
	sperm will evolve to oppose female meiotic drive.

@@

It seems that allowing sperm to influence the outcome of female meiosis would generate a confluence of these potential conflicts -- 
	sperm could actually assist an allele that distorts female meiosis.
However, this is not the case.
We find that an allele which acts through sperm to distort female meiosis in its favor %YB [too awkward / unclear?] 
	can rarely spread through a population if it bears any cost. 
Additionally, when this self-promoting driver can spread, it can only rarely 
	be maintained as a protected polymorphism, and due to its positive frequency dependence,  
	it spends very little time at intermediate frequency.
As such, this type of exploitation cannot generate a sustained genetic conflict.
It is therefore unlikely
	that female o\"{o}genesis and meiosis will evolve to prevent their effect.  
This likely explains why females can delay the completion of meiosis until after fertilization 
	without risking exploitation by collaborations between female
        drivers and sperm alleles.

Why is it that an allele that biases female meiosis in its favor can generate a genetic conflict, but an allele in sperm that assists this female driver cannot? 
So long as the transmission advantage of female meiotic drive outweigh the organismal fitness cost to heterozygotes, the female driver can spread when rare, and it increases in 		
	frequency until the fitness cost to homozygotes balances the transmission advantage.
By contrast, a sperm promoter of female drive is only effective when matched with a heterozygote female -- meaning that, when rare, this allele rarely enhances female drive. 
Even worse, when it does so it will preferentially find itself in a low fitness drive homozygote. 
Not only are drive-promoting sperm alleles unable to create a sustained genetic conflict, 
	but alleles in sperm with the opposite effect - that is those that prevent their own drive through female meiosis do maintain a polymorphism and 
	provide evolution  with time and opportunity to further minimize drive.
This is because such drive suppressing alleles reduced their chances
of forming low fitness homozygotes. 
More generally, natural selection favors alleles that act through sperm to
 reduce the opportunity of female meiotic drive regardless of linkage or phase.

@@

\section*{Appendix}
We highlight simple analytical results from: 
	the standard model of female meiotic drive (Model 1); a sperm-dependent 
	model of female meiotic drive (Model 2); a male-genotype-dependent model of
	female meiotic drive (Model 3); and more complex two-locus sperm-dependent drive models (Models 4-6). 
All models include a biallelic locus ($A/B$) with non-driving and driving alleles in frequencies $f_A$ and $f_B$, 
	respectively, while more complex models include more loci and/or alleles.  
Transmission rules describing the outcomes of all matings in each
	model are presented in a File S1 . 
The fitness of genotype, $g$ is sex-independent and equals $1$, $1-h_s$, and $1-s$ for genotypes $AA$, $AB$, and $BB$, respectively. %  and  $w_{AA} = 1$, $w_{AB} = 1-s_h$, and $w_{BB}=1-s$. 
Genotypic frequencies  equal  $f_g$ for adults in the current generation, 
	$f_g'$ in the next generation of zygotes (i.e. after recombination, random mating, drive, and syngamy, but before selection)
	and $f_g''$ in adults after selection. 
After a complete generation $f_g'' = f_g' w_g/\bar{w}$, where $\bar{w}$ is the population mean fitness and equals $\Sigma w_g f_g'$. 
Because many of our analyses assume genotype frequencies expected under Hardy Weinberg Equilibrium (HWE), we note which analytical results are approximations. 
We verify these approximations with exact numerical iterations in Figures S1-S3 and File S3 , and detail the derivation of these approximations in our File S2  .

\subsection*{Models 1-3. Single-locus drive}
\subsubsection*{Model 1. Traditional driver}
In the standard female drive model, meiosis in males is fair such that $A/B$ heterozygotes contribute $A$ and $B$ alleles with equal probabilities; however, $A/B$ females transmit the $B$ allele with probability $d>1/2$.  
We note that the timing of fertilization relative to female meiosis places another constraint on $d$, for example, if fertilization (and therefore, sperm dependent drive) takes place at MII (as in mammals),
	female drive requires an uneven number of crossovers between the centromere and the drive locus, 
	so $d$ is bounded to be $<0.83$ \citep[see ][ for discussion]{Buckler1999}. 
After drive and random mating, genotype frequencies are 
	$$f_{AA}'=f_A (f_{AA }+ f_{AB} (1 - d)),$$ 
	$$f_{AB}'=f_A (f_{BB} + f_{AB}d) + f_B (f_{AA} + f_{AB} (1 - d)),\text{ and}$$
	$$f_{BB}'= f_B (f_{AB} d + f_{BB}).$$ 
As detailed above, exact frequencies after drive, random mating and selection are $f_g''= f_g'w_g/\bar{w}$. 
Assuming HWE, a rare driver will spread when $s_h
        \lessapprox(2d-1)/(1+2d)$, and will fix when $s\lessapprox d -1/2 +
        3 s_h /2- d s_h$. This later inequality reduces to
        $s\lessapprox(2d-1)/2$ when the cost of drive is fully
        recessive.

\subsubsection*{Model 2. Single locus, sperm-dependent drive.}
Our single-locus model of sperm-dependent drive resembles the traditional driver, with the caveat that the $B$ allele drives in heterozygous females only when fertilized by $B$-bearing sperm. 
Therefore, genotype frequencies after drive are 
	$$f_{AA}'=f_A^2 , $$
	$$f_{AB}'=f_A f_B+f_B (f_{AA} + f_{AB}(1 - d)), \text{and} $$
	$$f_{BB}'= f_B (f_{AB} d + f_{BB} ).$$
%As above, exact frequencies after drive, random mating and selection are $f_i''= f_i'w_i/\bar{w}$. 
We iterate exact genotype frequency recursions ($f_g''= f_g'w_g/\bar{w}$) over generations to produce the frequency trajectories shown in the inset of 
Figure \ref{Eggsperm}B by plotting $f_B=f_{BB}+ \frac{1}{2}f_{AB}$
over time. %, and to assess the analytical approximations, below.
To assess invasion or fixation criteria, as well as bistability points, we iterate this system and
test whether $f_B$ increases over a grid of parameters.

\paragraph{Recessive fitness cost of self-promoting driver:} 
When fully recessive, the change in frequency of the self-promoting driver across generations equals 
\begin{equation}
\Delta f_B=f_A f_B^2 \left( [1-F][ d(1-2 f_A s) - f_B s-1/2] - s  F \right)/\bar{w}
\label{deltadriver}
\end{equation}
where $F$ is the deviation from genotypic frequencies expected under Hardy-Weinberg. 
Assuming HWE  ($F=0$) a common, recessive, self-promoting driver invades if $s\lessapprox(2 d - 1)/(4 d)$, and fixes if
 $s\lessapprox(2d-1)/2$. 
 Therefore, when 
\begin{equation}
(2 d - 1)/(4 d)  \lessapprox  s  \lessapprox  (2d-1)/2\label{ineqA}
\end{equation} 
a recessive, self-promoting driver will deterministically fix if drift,
 	mutation, or migration pressure bring its frequency above
\begin{equation} 
f^*_{B\text{ recessive}} \approx (1-2d+4ds)/(2s(2d-1)) \label{bistab_homozy_approx}
\end{equation}
 but it will be lost when introduced below this frequency. 
Compared to exact results (Figure S\ref{Bistab_homozyg_cost_fig}), Equations
 \eqref{ineqA} and \eqref{bistab_homozy_approx} offer reasonable, but imperfect approximations.
